A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network

### 2018-10-30 anzjs2018/sections/03_models.tex
more writing and pictures
edits

Due to the real-time nature of this application, we are using
a recursive Bayesian filtering approach as is often used in
real-time tracking applications.
We are using a separate model for each of the vehicle and network states,
and in each assume an underlying Markov process with state $\mathcal{X}_k$
with system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
of which observations $\mathcal{Y}_k$ are made
with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
giving the following general model:
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
The transition function $f$ describes the relationship between consecutive states
(i.e., what we want to model),
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

The following sections describe the two models used in this application.
The first, implemented using a particle filter, estimates vehicle states
to infer travel times along roads from a sequence of GPS positions;
the second uses these travel times to update road states 
and is implemented using an \emph{information filter},
a variant of the Kalman filter \citep{}.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance traveled $x$ in meters along the route and
it's speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are included in the state and estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We chose to employ a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

\begin{figure}[tb]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\;\;
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{The transition function $f$ models the behaviour of each particle,
            including changes in acceleration and stopping behaviour at bus stops.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
        \caption{Each particle's distance from the true observation is calculated using
            the geographic distance between two coordinates, which can then be used
            to calculate it's weight.}
        \label{fig:pf_state_update}
    \end{subfigure}\;\;
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{Resampling occurs using likelihood weights that use the distance
            between the particle's location and the vehicles reported location (red cross),
            resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{The vehicles state is estimated using a set of discrete points, or particles,
        each of which is independently mutated to make predictions of the future state. The history of the state estimates can be used to estimate parameters of interest, such as
        travel times along roads.}
    \label{fig:pf_state}
\end{figure}

@@

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size $N_{\text{eff}} = 1/(\sum_i (W_k^{(i)})^2)$
falls below a specified threshold $N_{\text{thres}}$.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_update}.
Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

The particle filter lets us to develop a model in which
the vehicle transition function $f$ implements several key bus behaviours:
\begin{itemize}
\item non-constant speed along roads,
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections or traffic lights (future work).
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's acceleration changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on the values used by \cite{Hans_2015};
future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.
% No intersection model is implemented here,
% as the segment road segments are between stops,
% and the travel times predicted by the network model at time $t_k$,
% $\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
% (see section~\ref{sec:kf}).
The transition function algorithm is shown in Figure~\ref{fig:algorithm},
which implements those features discussed above.

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection),
allow the placement of a bivate normal likelihood on the data
as shown in figure~\ref{fig:pf_lhood}.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
Now, the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is therefore calculated using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discretization of the state provided by the particle filter;
in other approaches, such as the Kalman filter,
the likelihood cannot be calculated using the distance as the estimate
is a distribution, and instead a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

@@

The primary objective of the network model is to estimate the \rt traffic conditions
(travel time) along roads in the transit network, 
as well as make short-term predictions for estimating arrival times.
In this paper,  we apply a very simple model in which each road is treated independently,
future work will investigate using correlations with adjacend roads and historical data
to improve the model and subsequent predictions.

@@

Travel times are assumed to be non-negative but otherwise normally distributed,
and constant over time---we do not attempt to model peak congestion in this paper.
Additionally the observations are non-constact and may be multiple as several buses travel
a road simultaneously.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state.

@@

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
The travel time estimates from the vehicle model for vehicle $m$ along segment $\ell$
is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
can be transformed to a measurement information covariance matrix 
$B_c^{m\ell} = (s^{m\ell}_c)^{-2}$
and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
The information update is now the sum of the information over all $M$ vehicles
that traversed segment $\ell$ since the last update, so
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}
\end{align*}
which are back transformed to obtain the desired parameter estimates
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}.
\end{equation*}

>>>>

Real-time tracking applications such as ours often use
a Bayesian filtering approach,
or recursive Bayesian estimation,
in which the data is processed sequentially instead of all at once,
reducing computational demands.
These models assume an underlying Markov process with state $\mathcal{X}_k$
and system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
of which observations $\mathcal{Y}_k$ are made
with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
giving the following general model:
\begin{equation}
\label{eq:rbe_model}
\begin{split}
\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
\end{split}
\end{equation}
The transition function $f$ describes the relationship between consecutive states,
while the measurement function $h$ is a deterministic function describing the
relationship between the underlying and observed states.

The following sections describe the two models used in this application,
both of which are examples of a Bayesian filter.
The first, implemented using a particle filter, estimates vehicle states
to infer travel times along roads from a sequence of GPS positions;
the second uses these travel times to update road states 
and is implemented using an \emph{information filter},
a variant of the Kalman filter \citep{Anderson_2012}.

@@

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance traveled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are included in the state and estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We chose to employ a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

\begin{figure}[tb]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\;\;
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{The transition function $f$ models the behaviour of each particle,
            including changes in acceleration and stopping behaviour at bus stops,
            which can result in multiple modes.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
        \caption{Each particle's distance from the true observation (red cross) 
            is calculated using
            the geographic distance between two coordinates, which can then be used
            to calculate it's weight.}
        \label{fig:pf_state_update}
    \end{subfigure}\;\;
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{Resampling occurs using likelihood weights that use the distance
            between the particle's location and the vehicles reported location (red cross),
            resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{The vehicles state is estimated using a set of discrete points, or particles,
        each of which is independently mutated to make predictions of the future state.
        Multimodality is easily handled, as demonstrated here,
        without degeneration of the filter.}
    \label{fig:pf_state}
\end{figure}

@@

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size 
\begin{equation}
\label{eq:neff}
N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
\end{equation}
falls below a specified threshold $N_{\text{thres}}$.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

The particle filter allows us to develop a model of several important bus behaviours:
\begin{itemize}
\item non-constant speed along roads,
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections or traffic lights (future work).
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's acceleration changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on the values used by \cite{Hans_2015};
future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.

The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
The $stop\_index$ function returns the index of a stop,
given a distance traveled and set of stop distances.
The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
which involves changing the vehicle's speed each second (i.e., acceleration).
Lines~13--25 handle arrival behaviour at stops:
if the particle passes the next stop,
it has a chance of stopping (lines~18--19),
in which case the dwell time is sampled and deducted from the 
time remaining to travel (lines~20--21).
Finally, to track travel time along segments,
at each point where the vehicle travels (lines 9 and 26),
the travel time for that segment, $z_j$, is incremented.
The mutation ends once the travel time $\delta$ reaches zero,
or when the vehicle reaches the end of the route (lines 10, 14).
Once finished, the particle's new state is returned (line 29).

\renewcommand{\algorithmicrequire}{\textbf{Start:}}
\newcommand{\algorithmicbreak}{\textbf{break}}
\newcommand{\Break}{\State \algorithmicbreak}
\begin{algorithm}[tb]
    \caption{Particle mutation function.}
    \label{fig:algorithm}
    \begin{algorithmic}[2]
    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
    Initialization $\delta \gets t_k - t_{k-1}$, 
        $j \gets stop\_index(x, \boldsymbol{S})$
    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
        \Comment{Remain longer at stop}
        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
        \State $\delta \gets \delta - \gamma - d_j$
    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
        \Comment{Occasionally bus stopped due to traffic}
        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
        \State $\delta \gets \delta - w$
        \State $z_j \gets z_j + w$
        \Comment{Congestion counts as travel time}
    \EndIf
    \While {$\delta > 0$ and $x < S_J$} 
        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
        \Comment{System noise, trunctated to be non-negative}
        \State $x \leftarrow x + \dot x$
        \If {$x \geq S_{j+1}$}
            \If {$j+1 = J$}
                \Comment{Break at end of route}
                \State $x \gets S_J$
                \Break
            \EndIf
            \State $u_j \sim \mathrm{U}(0,1)$
            \Comment{Simulate vehicle stopping}
            \If {$u_j < \pi$}
                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
                \State $\delta \gets \delta - \gamma - d_j$
                \State $x \gets S_j$
            \EndIf
            \State $j\gets j+1$
        \EndIf
        \State $\delta \gets \delta - 1$
        \State $z_j \gets z_j + 1$
        \Comment{Track travel time along road}
    \EndWhile
    \\
    \Return $x, \dot x, \boldsymbol z$
    \end{algorithmic}
\end{algorithm}

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
Now, the geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
cane be calculated using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discretization of the state provided by the particle filter;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

@@

The primary objective of the network model is to estimate the \rt traffic conditions
(travel time) along roads in the transit network, 
as well as make short-term predictions for estimating arrival times.
In this paper, we develop simple model in which each road is treated independently.
Future work will investigate using correlations with adjacent roads and historical data
to improve the model and subsequent predictions.

@@

Since multiple vehilces can travel along a road at the same time,
we have used an information filter to implement the network model.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state
in a single iteration.

@@

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
The travel time estimate of vehicle $m$ along segment $\ell$
is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
and is transformed to a measurement information covariance matrix 
$B_c^{m\ell} = (s^{m\ell}_c)^{-2}$
and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
The total information is the sum of the information over all $M$ vehicles
that traversed segment $\ell$ since the last update,
so the update equation is
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}.
\end{align*}
The desired parameter estimates (tavel time) are obtained 
by the inverse transformations
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}
\end{equation*}
which can now be used to predict the travel times of upcoming vehicles.

### 2018-10-30 anzjs2018/sections/04_implementation.tex
edits

\section{Real-time implementation and preliminary results}
\label{sec:rt}

@@

The general structure of the \rt component is:
\begin{enumerate}
\item Load GTFS data from database
\item Indefinitely repeat when new data recieved \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from new data
    \item Run particle filter on each vehicle to estimate update state
    \item Collect travel time information for each vehicle for any completed roads
    \item Update road network for any completed road segments
    \item Generate ETAs for vehicles uing combination of particle filter and network state
    \item Write ETAs to extended Google Protobuf binary file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarly concerned with ensuring each component of the program
is as efficient as possible,
allowing ETAs to be generated and distributed fast enough---our initial goal was 30~seconds.
Using \verb|C++| provides enough control over memory management
to ensure the program runs fast enough in \rt 
while still implementing the models.
We could also use OpenMP for parallelisation,
allowing us to process multiple vehicles simultaneously
depending on the number of cores available.

The particle filter requires thousands of particles per vehicle,
so it is necessary to explore the performance of the application
with varying number of particles.
It is also necessary to assess the performance of the particle filter,
and compare values for the fixed model parameters.
To do so, we implemented a simulated \rt environment
in which we could analyse the same subset of real vehicle data from 8~October, 2018
using a range of settings.
These simulations were carried out of a Virtual Machine 
with 8~cores and 32~Gb of memory, 
running Ubuntu~16.04 and using R~3.4.1.

\begin{figure}[tb]
    \centering
    % \includegraphics[width=0.7\textwidth]{figures/04_model_results_gps.pdf}
    \caption{The distance between the observation and the nearest point on the shape.}
    \label{fig:gps_dist}
\end{figure}

@@

During development of our application,
our primary concert was the computational faesibility of a particle filter.
For each iteration, 
the timings of the individual components in 2a--f above are recorded.
Since the number of vehicles traveling at any given time changes throughout the day,
we used the average timings over a 15~minute window to compare.

@@

The model was evaluated using several statistics,
each calculated using a range of parameter values.
The first is effective sample size,
defined in (\ref{eq:neff}),
which is used to determined if resampling is required.
Another important statistic is the \emph{degeneration rate},
or the percent of iterations where the particle filter
is unable to find any plausible states given the observation,
and the vehicle is reinitialised.
Note, however, that the lower limit of this value 
is attributed to \emph{invalid trips},
however there is no simple way of filtering these trips.
The last statistic we compute is the variance of road segment travel times
over time.
For this example, we varied the number of particles $N$,
GPS error $\epsilon$, and system noise $\sigma^2$.

The one parameter we can determine from the data is GPS error, $\epsilon^2$,
by looking at the distance between observations and the shape.
Figure~\ref{fig:dist_to_route} shows the distribution of distance between the vehicle
observation and the \emph{nearest point on the path}.
Due to the way the observations are reported, 
many observations will be ``exactly'' on the route at a bus stop,
which explains the spike at just above zero~meters.

@@

Lastly is the variability of travel times and their uncertainties.
These are shown in Figure~\ref{fig:travel_times}.
System noise again shows no effect on the variability of travel time estimates,
however increasing GPS error tends to increase travel time variability.
This means that there is a trade-off between the models performance
and how well it estimates the main parameter of interest:
larger values of $\epsilon$ result in less degeneration and larger $N_\text{eff}$,
at the cost of more variable travel time estimates.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_times.pdf}
    \caption{Travel times.}
    \label{fig:travel_times}
\end{figure}

>>>>

\section{Real-time implementation}
\label{sec:rt}

@@

The general structure of the \rt component is shown below,
with the bold steps being those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Indefinitely repeat when new data recieved \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from new data
    \item \textbf{Run particle filter on each vehicle to update state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles uing combination of particle filter and network state
    \item Write ETAs to extended Google Protobuf binary file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarly concerned with ensuring each component of the program
is as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be faesible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provides enough control over memory management
to ensure the program runs fast enough in \rt 
while still implementing the models.
We also use OpenMP for parallelisation,
allowing us to process multiple vehicles simultaneously,
which are each independent so threadsafety is a non-issue.

The particle filter requires thousands of particles per vehicle,
so it is necessary to explore the performance of the application
with varying number of particles.
We also assess the performance of the particle filter
for a range of fixed model parameters values,
such as system noise and GPS error.
To do so, we implemented a simulated \rt environment
in which we could analyse the same subset of real vehicle data from 8~October, 2018
using a range of settings.
These simulations were carried out of a Virtual Machine 
with 8~cores and 32~Gb of memory, 
running Ubuntu~16.04 and using R~3.4.1.

% \begin{figure}[tb]
%     \centering
%     % \includegraphics[width=0.7\textwidth]{figures/04_model_results_gps.pdf}
%     \caption{The distance between the observation and the nearest point on the shape.}
%     \label{fig:gps_dist}
% \end{figure}

@@

During development of our application,
our primary concern was the computational faesibility of a particle filter.
For each iteration, 
the timings of the individual components are recorded.
Since the number of vehicles traveling at any given time changes throughout the day,
we used the average timings over a 15~minute window to compare.

@@

The model was evaluated using several statistics,
each calculated using a range of parameter values.
The first is effective sample size,
defined in (\ref{eq:neff}),
which is used to determined if resampling is required.
Another important statistic is the \emph{degeneration rate},
or the percent of iterations where the particle filter
is unable to find any plausible states given the observation,
and the vehicle is reinitialised.
Note, however, that some observations are \emph{invalid trips},
in which the vehicle is logged in to the wrong trip,
or ``off duty'' while still transmitting for a trip.
Some of these can be filtered for being excessively far from the route,
but others---for example traveling the wrong direction---cannot be.
The last statistic we compute is the variance of road segment travel times
over time.
For this example, we varied the number of particles $N$,
GPS error $\epsilon$, and system noise $\sigma^2$.

The one parameter we can investigate from the data is GPS error, $\epsilon^2$,
by looking at the distance between observations and the shape.
Figure~\ref{fig:dist_to_route} shows the distribution of distance between the vehicle
observation and the \emph{nearest point on the path}.
We see two peaks,
one at about 1m, and another at 2.5m.
The first of these is possibly an artefact of the data,
for example automated waypoints at intersections.
We have explored a range of GPS error values from 1 to 5~m.

@@

Lastly is the uncertainty of travel time estimates,
as shown in Figure~\ref{fig:travel_times}.
System noise again shows no effect on the variability of travel time estimates,
however increasing GPS error tends to increase travel time variability.
This means that there is a trade-off between the models performance
and how well it estimates the main parameter of interest:
larger values of $\epsilon$ result in less degeneration and larger $N_\text{eff}$,
at the cost of more variable travel time estimates.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_times.pdf}
    \caption{The within-segment posterior travel time uncertainty (standard error), subset by number of particles.}
    \label{fig:travel_times}
\end{figure}

### 2018-10-30 anzjs2018/sections/06_discuss.tex
edits

The main goal of this work was to explore the faesibility
of a particle filter as a \rt estimation method for transit vehicle states,
and the use of estimated states to determine traffic conditions
throughout the network.
This allows vehicles to inform others---irrespective of route---of 
how long it will take to reach future stops.

Not only is the particle filter a viable choice in terms of speed,
it has shown it can handle most situations without losing the vehicle,
with the exception of when the vehicle is not traveling along
the specified route.
Hopefully with future development,
the stopping behaviour can be more accurately modeled,
perhaps incorporating trip updates into the likelihood function.

The network model is still in the early stages of development,
but given that the implementation is running quickly---twice as fast as our target---%
there is room for developing a more complex model.
Future work will explore the use of historical data to develop informative priors
for travel times,
which will provide a useful fallback in the case of no data,
or for making more accurate long-term predictions.
Once such a model has been developed, we plan to explore prediction estimates,
particuarly comparing point and intervals as a means of communicating 
uncertainty to commuters.

>>>>

The main goal of this work was to explore the faesibility
of a particle filter as a \rt estimation method for transit vehicle states,
and the use of estimated states to determine traffic conditions
throughout the network.
This allows vehicles to inform others---irrespective of route---of 
current \rt traffic conditions,
and provide a means of obtaining more reliable ETAs.

We have shown that the particle filter a viable \rt option in terms of speed,
and while it occasionally loses the bus,
a lot of this is attributed to invalid data (incorrect route, direction, etc).
Future development will incorporate more accurate stopping behaviour,
perhaps using GTFS arrival and departure times in the likelihood function.

The network model is still in the early stages of development,
but given that the implementation is running well within our 30~second target,
there is room for developing a more complex model.
Future work will explore the use of historical data to develop informative priors
for travel times,
which will provide a useful fallback in the case of no data,
or for making more accurate long-term predictions.
Once such a model has been developed, we plan to explore prediction estimates,
particuarly comparing point and intervals as a means of communicating 
uncertainty to commuters.
